Machine learning methods in streamflow prediction considering climate variable windowing

### 2021-06-28 paper/paper.tex
Sobre o gráfico de comparação no paper

To evaluate the performance of SVR, RF, Stacking Regressor and decision tree, we use two criteria: Mean Absolute Error (MAE) and R-squared. MAE represents the average of the absolute difference between the actual and predicted values. We included MAE as an error metric because it provides an interpretation of error on the same scale of streamflow. This procedure is represented by the Equation
    \begin{equation}
        \label{eqn:mae}
        MAE={\frac{1}{n}\sum_{i=0}^{n-1}|(X_{pred}-X_{i})|}
    \end{equation}
in which $n$ means the number of non-missing data, $i$ is the variable, $X_{pred}$ and $X_{i}$ mean the predicted and the actual value, respectively. Conversely, $R^2$ represents the proportion of the variance in the dependent variable which is explained by the regression model, given by the Equation
    \begin{equation}
        \label{eqn:r2}
        R^2={1 - \frac{\sum\limits_{i=0}^{n-1}(X_{pred}-X_{i})^{2}}
                      {\sum\limits_{i=0}^{n-1}(X_{i}-X_{mean})^{2}}}
    \end{equation}
where $X_{mean}$ is $\frac{1}{n}\sum_{i=0}^{n-1}X_{i}$ and $X_{pred}$ and $X_{i}$ are the predicted and it's corresponding true value, while $n$ is the size of the test sample. R-square is a indicator of goodness, indicating how well the model can predict unseen samples. The best possible R-squared value is $1.0$ and it can be negative, as the model gets arbitrarily worse.

@@

For month variable we apply one-hot encoding, which is often used for categorical variables.

@@

Having tested horizon prediction of streamflow and water level varying from 1 to 29 days, and windowing at intervals of 1, 15 and 30 days for the climate variables, results indicated better fit for horizon of one day and window of 30 days. However, the models presented different results of R\textsuperscript{2} when they are compared to each other, as shown in Figures \ref{fig:r2_1}, \ref{fig:r2_15} and \ref{fig:r2_30}.

@@

@@

SVR had the highest values of MAE, while RF had the lowest values. The Stacking Regressor and decision tree produced similar values of MAE. Surprisingly, RF presented a slight increase of MAE for greater days of lagging, and for the case of 15 and 30 days-windowing, while the other models presented a trend of having a decline.

@@

>>>>

To evaluate the performance of SVR, RF, Stacking Regressor and decision tree, we use two criteria: Mean Absolute Error (MAE) and R-squared. MAE represents the average of the absolute difference between the actual and predicted values. We included MAE as an error metric because it provides an interpretation of error on the same scale of streamflow. This procedure is represented by the Equation
    \begin{equation}
        \label{eqn:mae}
        MAE={\frac{1}{n}\sum_{i=0}^{n-1}|(X_{pred}-X_{i})|}
    \end{equation}
in which $n$ means the number of non-missing data, $i$ is the variable, $X_{pred}$ and $X_{i}$ mean the predicted and the actual value, respectively. Conversely, $R^2$ represents the proportion of the variance in the dependent variable which is explained by the regression model, and is given by the Equation
    \begin{equation}
        \label{eqn:r2}
        R^2={1 - \frac{\sum\limits_{i=0}^{n-1}(X_{pred}-X_{i})^{2}}
                      {\sum\limits_{i=0}^{n-1}(X_{i}-X_{mean})^{2}}}
    \end{equation}
where $X_{mean}$ is $\frac{1}{n}\sum_{i=0}^{n-1}X_{i}$, and $X_{pred}$ and $X_{i}$ are the predicted and it's corresponding true value, while $n$ is the size of the test sample. R-square is a indicator of goodness, indicating how well the model can predict unseen samples. The best possible R-squared value is $1.0$ and it can be negative, as the model gets arbitrarily worse.

@@

For the month variable, we applied one-hot encoding, which is often used for categorical variables, when the regressor was sensitive to the scale of the attributes, such as SVR, or Ordinal encoding, otherwise.

@@

Having tested horizon prediction of streamflow and water level varying from 1 to 29 days, and windowing at intervals of 1, 15 and 30 days for the climate variables, results indicated better fit for horizon of one day and window of 30 days. However, the models presented different results of R\textsuperscript{2} when they are compared to each other, as shown in Figure~\ref{fig:r2}.

@@

From an evaluation of Figure~\ref{fig:StackedComparisson}, which demonstrates the true versus predicted values for 1 and 29 days of prediction horizon, it is evident that the model trained for 1 day of horizon presents greater sensitivity to sudden changes in reservoir streamflow, while that the model with the longest horizon time, despite being able to follow the streamflow trend, does not faithfully represent the peaks present. Another point that can be justified by the presence of these peaks is the relatively large value found for the MAE since the model ends up being highly penalized by sudden changes in streamflow. On the other hand, an R-squared metric, which considers the amplitude of the actual data, ends up not being penalized in the same way. Figure~\ref{fig:StackedComparisson} presents data for a two-year window from 2000 to 2002, so data can be more easily visualized. For a visualization for the whole data range, see Appendix~\ref{sec:TrueVersusPredicted}

\begin{figure}[htbp]
    \centering
    \includesvg{Graphs/StackedComparisson}
    \caption{Comparison between true and predicted values for a prediction     horizon of 1 and 29 days. Visualization for a 2 year window, from 2000 to 2002}
    \label{fig:StackedComparisson}
\end{figure}

@@

SVR had the highest values of MAE, while RF had the lowest values. The Stacking Regressor and decision tree produced similar values of MAE. Surprisingly, RF presented a slight increase of MAE for increased prediction horizon, and for the case of 15 and 30 days-windowing, while the other models presented a trend of having a decline.

@@

As the prediction horizon increases, the tested models become less sensitive to drastic changes in the streamflow, not being able to predict peaks, but keeping the ability to describe the overall tendency.
